---
title: "MPG difference between transmissions types"
output: pdf_document
geometry: margin=0.5in
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=2.5, comment="",
                      echo=FALSE, warning=FALSE, message=FALSE)
```
```{r load_R_file, cache=FALSE}
knitr::read_chunk('project.R')
```

## Executive summary

This report presents results on a problem how transmission type of a car
influence on its miles per US gallon (MPG). To explain this relationship, 
a robust model with transmission type and weight variables is presented.
A car with manual transmission has bigger initial MPG (by value of almost 15), 
but with the increase of its weight, the final MPG decreases almost 3 times 
more than in case automatic transmission.

## Introduction

This report investigates the relationship between a set of variables and miles 
per US gallon. In particular, the report focuses on two main issues: 1) is an 
automatic or manual transmission better for MPG? 2) quantification of the MPG 
difference between automatic and manual transmissions.

The studied data was extracted from the 1974 Motor Trend US magazine, and 
comprises fuel consumption and 10 aspects of automobile design and performance 
for 32 automobiles (1973–74 models)[^1]. The data consists of 32 observations 
on 11 variables: *mpg* - miles/(US) gallon; *cyl* - number of cylinders; *disp* - displacement (cu.in.); *hp* - gross horsepower; *drat* - rear axle ratio; *wt* - weight (lb/1000); *qsec* - 1/4 mile time; *vs* - V/S; *am* - transmission (automatic or manual); *gear* - number of forward gears; *carb* - number of carburetors.
 
[^1]: Henderson and Velleman (1981), Building multiple regression models 
interactively. *Biometrics*, **37**, 391–411.

```{r setup}
```
```{r libraries}
```
```{r set-strings}
```
```{r preprocessing}
```

```{r}
library(knitr)
```

## Exploratory data analysis

```{r exploratory-data-analysis}
```

A brief look at the data shows that there is a difference between MPG within 
transmission groups. The unpaired $t$-test for the means unequality in 
automatic and manual groups has `r round(t_result$p.value, 3)` $p$-value.

The following table shows values of Pearson correlation between MPG and 
the other variables, for both types of the transmission and for a given type.
The number of cylinders, the displacement, the gross horsepower and the weight
are highly correlated to MPG (absolute values greater than $0.75$), considered 
jointly as well as separated. Such variables will be deeply investigated in the 
next step. A visualization of the relatioship between MPG and the variables
is in the **Appendix** section.

```{r}
kable(round(mtcData.corr[,c("cyl","disp","hp","drat","wt","qsec","vs","gear",
                            "carb")], 2),
      caption="Pearson correlations between MPG and the other variables 
      (columns) with respect to the types of the transmission (rows).")
```

## Fitting multiple models

```{r fit-multiple-models}
```

The variables mentioned in **Exploratory data analysis** section where selected 
as potential regressors. Several linear models with combinations of the 
variables were investigated. Only variants of variable-transsmision type 
interactions were considered. After the elimination process remained only the 
models which: have overall $p$-values less than $0.05$, each coefficient
has also $p$-value less than $0.05$, have reasonably high adjusted $R^2$ values
(around $0.75$ and more) and have arbitrary simple formula.
The following models were selected to further evaluation (adjusted $R^2$ values 
in parentheses):

 1. `mpg ~ wt:am          ` (`r round(rsq_vals[1], 3)`),
 2. `mpg ~ wt:am + am     ` (`r round(rsq_vals[2], 3)`),
 3. `mpg ~ wt:am + hp:am  ` (`r round(rsq_vals[3], 3)`),
 4. `mpg ~ disp:am + am   ` (`r round(rsq_vals[4], 3)`),
 5. `mpg ~ hp:am          ` (`r round(rsq_vals[5], 3)`).

## Final model selection

```{r model-selection}
```

ANOVA tables were computed for nested models: 1. vs 2., 1. vs 3. and 3. vs 5.
All significantly differ. The $p$-values are 
`r round(model_comp[[1]][["Pr(>F)"]][2], 3)`, 
`r round(model_comp[[2]][["Pr(>F)"]][2], 3)` and
`r round(model_comp[[3]][["Pr(>F)"]][2], 3)`, respectively.

As the final model was chosen a model with the highest adjusted $R^2$ value, 
which is model `r which.max(rsq_vals)`. with the `r round(max(rsq_vals), 3)` 
value.

### Coefficients interpretation

```{r coefficients-interpretation}
```

Coefficients of the final model are as follow:

```{r coefficient-display}
round(coeff, 2)
```

An interpreation of the coefficients is following:

 1. we treat `r abs(round(coeff, 2)[1])` as the starting MPG of a car without
 knowledge about its weight and transmission type parameters,
 2. we estimate an expected `r abs(round(coeff, 2)[2])` increase in MPG in case of
 manual transmission in holding remaining variables constant,
 3. we estimate an expected `r abs(round(coeff, 2)[3])` decrease in MPG for every
 lb/1000 increase in weight of a car with an automatic transmission in holding 
 the remaining variables constant,
 4. we estimate an expected `r abs(round(coeff, 2)[4])` decrease in MPG for every
 lb/1000 increase in weight of a car with a manual transmission in holding 
 the remaining variables constant.

### Residual plots and diagnostics

```{r residual-plots-and-diagnostics}
```

The residual plot (see **Appendix** section) presents relation between predicted
values and residuals. As it can be seen, the model is unbiased and 
homoscedastic. Studentized Breusch-Pagan test also shows that there is no 
evidence to suspect heteroscedasticity ($p$-value `r round(bp_pval, 3)`).

The most influential instance in the dataset is considered. The change
in the predicted response when `r rownames(mtcData)[changer]` is deleted 
in fitting the model is `r round(changer_fits, 2)`. The change in individual 
coefficients when the car is deleted in fitting the model is listed below:

```{r changer-dfbetas}
round(changer_betas, 3)
```

### Uncertainty quantification

```{r uncertainty-quantification}
```

To assess the precision of the predictions, the residual standard error is used.
It represents the average distance that the observed values fall from the
regression line. In that case it is `r round(resid_std_err, 2)`, as it is an
uncertainty measurement of the outcome MPG.

### Transmission difference quantification

```{r transmission-diff-quantification}
```

Since it is hard to quantify the difference between transmission types with
respect to MPG and other variables, some basic indicators are used. As it was 
mentioned in **Introduction** section, the difference of MPG averages between 
transmission groups is statistically significant. The mean MPG for automatic 
transmission equals to `r round(mpg_mean.am0, 2)` and for manual transmission 
equals to `r round(mpg_mean.am1, 2)`.

## Summary

In general, manual transmission is better for MPG. To understand what influences
on MPG, the model with transmission type and weight variables was considered.
When we investigate the parameters of the cars, that one with manual 
transmission has bigger initial MPG, but the increase of its weight decreases 
final MPG almost 3 times more than in case automatic transmission.

\newpage

## Appendix

### Exploratory data analysis plots

```{r plot-am}
print(expl_plots.am)
```

```{r plots-cyl}
print(expl_plots.cyl)
```

```{r plots-disp}
print(expl_plots.disp)
```

```{r plots-hp}
print(expl_plots.hp)
```

```{r plots-wt}
print(expl_plots.wt)
```

### Residuals plot

```{r residual-plot}
print(resid_plot)
```