---
title: 'Analysis of the ToothGrowth data'
output: pdf_document
---

In this paper we will investigate ToothGrowth data (see [dataset documentation]). The data contains the response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid). A data frame with 60 observations on 3 variables:

 * *len* (numeric) - tooth length,
 * *supp* (factor) - supplement type (VC or OJ),
 * *dose* (numeric) - dose in milligrams.

We will compare tooth growth with supplement type and dose by hypothesis tests.

# Initial setup

Before we start data processing, we load necessary libraries for plots.

```{r, results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
```

# Loading and preprocessing the data

We perform a basic summary of the data.

```{r}
data(ToothGrowth)
str(ToothGrowth)
print(summary(ToothGrowth))
```

As we can see, we have 3 attributes. The *dose* attribute is numeric but for analysis purpose it is better to turn it into a factor variable. The attribute has only 3 levels.

```{r}
ToothGrowth$dose = factor(ToothGrowth$dose)
```

Let us perform some basic exploratory data analyses.

```{r, fig.width=7, fig.height=3.5, results='hide', message=FALSE, warning=FALSE}
print(ggpairs(data=ToothGrowth,
              upper = list(continuous = "points",
                           combo      = "dot",
                           discrete   = "facetbar"),
              lower = list(continuous = "cor",
                           combo      = "facethist",
                           discrete   = "ratio"),
              colour="dose"))
```

As one can see, we have have the same proportions of data either we split it by *supp* or *dose* attribute. When we split the data by *supp*, it is hard to tell whether the means are the same. But when we split the data by *dose*, the average lenght of 3 groups seem to differ. We will focus on this problem in the next section.

# Comparison of tooth growth by supp and dose
 
We use confidence intervals and hypothesis tests to compare tooth growth by supp and dose attributes. We perform paired t-tests on equality of means with 95% confidence level. Since the data contains results for 10 guinea pigs in each *supp* and *dose* combination, the t-test is paired. We can not say anyting about equality of variances, so we assume that they are not equal. For each test a null hypothesis $H_0$ is that a difference in means is equal to 0; an alternative hypothesis $H_a$ is that true difference in means is not equal to 0.

Firstly, we perform t-test when we split data by *supp* attribute.

```{r}
TG_list = split(ToothGrowth, ToothGrowth$supp)
TGsOJ = TG_list[[1]]
TGsVC = TG_list[[2]]

t_s = t.test(x = TGsOJ$len, y = TGsVC$len,
             alternative = c("two.sided"),
             mu = 0, paired = TRUE, var.equal = FALSE,
             conf.level = 0.95)
```

Secondly, we perform t-tests when we split data by *dose* attribute.

```{r}
TG_list = split(ToothGrowth, ToothGrowth$dose)
TGd05 = TG_list[[1]]
TGd10 = TG_list[[2]]
TGd20 = TG_list[[3]]

t_d0510 = t.test(x = TGd05$len, y = TGd10$len,
                 alternative = c("two.sided"),
                 mu = 0, paired = TRUE, var.equal = FALSE,
                 conf.level = 0.95)
t_d0520 = t.test(x = TGd05$len, y = TGd20$len,
                 alternative = c("two.sided"),
                 mu = 0, paired = TRUE, var.equal = FALSE,
                 conf.level = 0.95)
t_d1020 = t.test(x = TGd10$len, y = TGd20$len,
                 alternative = c("two.sided"),
                 mu = 0, paired = TRUE, var.equal = FALSE,
                 conf.level = 0.95)
```

# Results and conclusions

Let us look on the result of the t-test in by-*supp* split groups:

```{r}
print(c(t_s$conf.int[1:2], t_s$p.value))
```

The interval is entirely above zero. The p-value is lower than 0.05. 
There is sufficient evidence to reject $H_0$, so we reject $H_0$ in favor of $H_a$.

Now let us look on the results of the t-tests in by-*dose* split groups:

```{r}
print(c(t_d0510$conf.int[1:2], t_d0510$p.value))
```
```{r}
print(c(t_d0520$conf.int[1:2], t_d0520$p.value))
```
```{r}
print(c(t_d1020$conf.int[1:2], t_d1020$p.value))
```

All intervals are entirely below zero. In each case p-value is lower than 0.05.
In each case there is sufficient evidence to reject $H_0$, so in each case we 
reject $H_0$ in favor of $H_a$.

To conclude, the interpreation of results is that the length of teeth depends on dose level of Vitamin C and on a delivery method.

[dataset documentation]:https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html